Revisiting the Bias-Variance Tradeoff of Multi-Task Learning in High Dimensions

### 2020-08-09 neurips20_response/response.tex

\documentclass{article}

\usepackage{neurips_2020_author_response}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{wrapfig}
\usepackage{graphicx}
\usepackage{enumitem}

\usepackage{xcolor}
\newcommand{\yell}[1]{\textcolor{red}{#1}}
\newcommand{\todo}[1]{\textcolor{red}{{TODO: #1}}}
\newcommand{\num}[1]{\textcolor{black}{#1}}

% Squished list environment to save space
\newcommand{\squishlist}{
\begin{list}{$\bullet$}
{ \setlength{\itemsep}{0pt}
\setlength{\parsep}{1pt}
\setlength{\topsep}{1pt}
\setlength{\partopsep}{0pt}
% \setlength{\labelwidth}{1em}
}
}
\newcommand{\squishend}{\end{list}}

\begin{document}

We thank all of our reviewers for carefully reading our work and providing insightful reviews.
First, we address several main issues that are raised by the reviewers.
Then, we respond to each individual reviewer's feedback.

\textbf{Motivation for our setting. [R1, R2, R3]}
As R2 pointed out, we focus on a situation where we have a target task for which we only have limited labeled data and several source tasks.
We study when training the tasks together can benefit the target task.
While this setting is different from traditional MTL, it is commonly encountered in practice. For example, the task of interest might be about predicting a rare event or classifying an Xray-scan.
For such settings, collecting large amounts of labeled data for the task is either not possible or very expensive, but auxiliary labeled data are often easier to obtain.
Traditional MTL theory that studies the average performance of all tasks does not help predict when adding the source tasks provides positive transfer for the target task or not.
Our work applies to these settings and takes a step towards filling the gap.

\textbf{Extension to multiple tasks.}
We thank R1 and R2 for raising the important question of whether our setting applies to multiple tasks.
First, for multi-label settings where all tasks have the same features, i.e. $X_i = X$ for any $i$, our result in Theorem 3.6 provides a sharp bias-variance tradeoff of MTL and \textit{all of our insight for two tasks except covariate shift applies to multi-label settings}.
Due to page limit we have not elaborated on this part in the submission, but we will elaborate the theoretical implication in the next version.
The reason why the issue about covariate shift does not apply to this setting is because the tasks have the same features, hence there is no covariate shift.

Second, for settings where different tasks have different features, we can still extend our qualitative insight here.
That is, we can show that \textit{as long as the dimension of the shared layer ($B$) is smaller than the total number of tasks, the variance of the MTL estimator for the target task is always smaller than the variance of the STL estimator but the bias is always larger}.
We will include this result in the next version.

\textbf{Clarifying writing. [R2, R3]}
We thank R2 and R3 for providing numerous detailed feedback that help us improve writing.

\textbf{R1:}

\textbf{R2:}

\textbf{R3:}

\end{document}

### 2020-08-09 arxiv/app_exp.tex
author response

\subsection{Synthetic Settings}\label{app_synthetic}
	We describe the parameters used for producing Figure \ref{fig_model_shift_phasetrans}.
	For all three simulations, we use dimension $p = 200$.
	In Figure \ref{fig_model_shift}, we use $\rho_1 = 90, \rho_2 = 30$ for sample sizes.
	The noise level is $\sigma = 5$.
	We set $\kappa = 1$ and vary model distance $d$ from $0.01$ to $0.2$.
	The $x$-axis measure the function $\Phi(\beta_1, \beta_2)$ described in Section \ref{sec_similarity}.
	In Figure \ref{fig_size}, we set $\kappa = 1$ and $d = 0.02$ for model distance.
	We use $\rho_2 = 500$ and vary $\rho_1$ from $50$ to $800$ for sample sizes.
	In Figure \ref{fig_covariate}, we set $\kappa = 1$ and $d = 0$.
	We set $\rho_2 = 4$ and vary $\rho_1$ from $5$ to $25$ for sample sizes.
	We use the scale parameter $\lambda = 1$ for the curve without covariate shift and $\lambda = 2$ for the curve with covariate shift (cf. Section \ref{sec_covshift}).
	Without loss of generality, we have rescaled the $x$ and $y$ axis for the ease of presentation.
	This can be achieved similarly by rescaling the task data.

@@

For the text classification experiment, we encode each word using the GLoVe word vectors.
\footnote{http://nlp.stanford.edu/data/wordvecs/glove.6B.zip}
We apply an average pooling layer over the word embeddings.
Then we add a multi-layer perception (MLP), LSTM, or CNN layer on top of the word embeddings \cite{lei2018simple}.

\textbf{Single-task based metric.}
We simplify the setup of this experiment by fixing the sample sizes of every task pair.
For sentiment analysis tasks, the sample size of the source task ranges in $500, 1000, 1500$.
We randomly sample these many data points from the task.
The sample size of the target task is $1000$.
For image classification tasks, the training sample size is $10,000$ for every task.

\textbf{Incremental training schedule.}
For the case of two tasks, we compare the incremental training scheduled described in Algorithm \ref{alg_inc_train} below to a round-robin training schedule baseline.
We add $20\%$ of source task samples for every $T = 2$ epochs.
The total number of epochs is $20$.
We set the threshold $\tau$ as the accuracy of the target task using the baseline training schedule.

@@

\textbf{Theoretical validation.}
For Figure \ref{fig_ab_sim}, we select a similar and a dissimilar source task compared to the target task using domain knowledge.
First pair: the customer review dataset (CR) , which predicts whether a review is positive or negative, is more similar to SST (sentiment treebank) than MPQA (question type).
Second pair: SST is more similar to MR since they both concern about positive or negative opinions expressed the text.
TREC is less similar to MR because the task is about question types.
Third pair: MPQA (opinion polarity) is more similar to TREC (question type)

For Figure \ref{fig_ab_data} and Figure \ref{fig_ab_cov}, we fix the sample size of the target task as $1000$.
We vary the sample size of the source task from $100$ to $3000$.
For Figure \ref{fig_ab_cov}, we implement the covariance alignment procedure following \cite{WZR20}.

@@

\textbf{Further results of the covariance alignment procedure.}
Our results in Figure \ref{fig_ab_cov} are averaged over all the task pairs.
In Figure \ref{fig_covariate_app}, we show two task pairs as examples.
In Figure \ref{fig_cov_a}, we observe that for the particular task pair, covariance alignment provides more significant gains when the sample ratio is large.
In Figure \ref{fig_cov_b}, we observe that covariance alignment does not always help.
One explanation is over-fitting because of additional parameters from the alignment module.
An interesting question is to understand when adding the alignment module benefits the multi-task learning model.
We leave this question for future work.
%Note: For text classification tasks, the source task training data size ranges from 500 to 1,500 and target task training data size is 1000; For ChestX-ray14,

\begin{figure}%[!t]
	\centering
	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\includegraphics[width=0.75\textwidth]{figures/ratio_alignment_norm_trec_cr_lstm.pdf}
		\caption{Task pair TREC and CR}
		\label{fig_cov_a}
	\end{subfigure}\hfill
		\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\includegraphics[width=0.75\textwidth]{figures/ratio_alignment_mr_sst_lstm.pdf}
		\caption{Task pair MR and SST}
			\label{fig_cov_b}
	\end{subfigure}
	\caption{(a) For the task pair TREC and CR, adding the covariance alignment procedure provides more improvement when the source/target sample ratio is large.
	(b) For the task pair MR and SST, adding the covariance alignment procedure hurts performance.
	We suspect that this happens because of over-fitting on the extra parameters in the additional covariance alignment module.}
	\label{fig_covariate_app}
\end{figure}

>>>>

\subsection{Synthetic Settings}\label{app_synthetic}
	We describe the parameter settings used in the synthetic experiments (Figure \ref{fig_model_shift_phasetrans}).
	For all three synthetic experiments, we set the input feature dimension as $p = 200$.
	\squishlist
		\item Task similarity: We use $\rho_1 = 90, \rho_2 = 30$ for sample sizes.
	The noise level is $\sigma = 5$.
	We set $\kappa = 1$ and vary model distance $d$ from $0.01$ to $0.2$.
	The $x$-axis measures the function $\Phi(\beta_1, \beta_2)$ described in Section \ref{sec_similarity}.
	The $y$-axis measures the improvement of prediction loss using MTL compared to STL.
		\item Sample size: We set $\kappa = 1$ and $d = 0.02$ for model distance.
	We use $\rho_2 = 500$ and vary $\rho_1$ from $50$ to $800$ for sample sizes.
		\item Covariate shift: We set $\kappa = 1$ and $d = 0$.
	We set $\rho_2 = 4$ and vary $\rho_1$ from $5$ to $25$ for sample sizes.
	We use the scale parameter $\lambda = 1$ for the curve without covariate shift and $\lambda = 2$ for the curve with covariate shift (cf. Section \ref{sec_covshift}).
	Without loss of generality, we have rescaled the $x$ and $y$ axis for the ease of presentation.
	This can be achieved similarly by rescaling the task data.
	\squishend

@@

For the text classification experiment, we encode each word using the GLoVe word embeddings.%
\footnote{http://nlp.stanford.edu/data/wordvecs/glove.6B.zip}
We evaluate three model choices.
For multi-layer perception (MLP), we apply an average pooling layer over the word embeddings.
For LSTM and CNN, we add a shared feature representation layer on top of the word embeddings \cite{lei2018simple}.

\textit{Predicting transfer effect via STL results.}
We simplify the setup of this experiment by fixing the sample sizes of every task pair.
For sentiment analysis tasks, the sample size of the source task ranges in $500, 1000, 1500$.
We randomly sample these many data points from the task.
The sample size of the target task is $1000$.
For image classification tasks, the training sample size is $10,000$ for every task.

\textit{Mitigating negative transfer via incremental training.}
For the case of two tasks, we compare the incremental training scheduled described in Algorithm \ref{alg_inc_train}  to a round-robin training schedule baseline.
We add $20\%$ of source task samples for every $T = 2$ epochs.
The total number of epochs is $20$.
We set the threshold $\tau$ as the validation accuracy of the target task using the baseline training schedule.

@@

\textit{Validating the Theoretical Results.}
We fill in the details of the experimental procedure used for the results in Figure \ref{fig_ablation}.
\squishlist
	\item Task similarity: We select a similar and a dissimilar source task compared to the target task using domain knowledge.
First pair: the customer review dataset (CR) , which predicts whether a review is positive or negative, is more similar to SST (sentiment treebank) than MPQA (question type).
Second pair: SST is more similar to MR since they both concern about positive or negative opinions expressed the text.
TREC is less similar to MR because the task is about question types.
Third pair: MPQA (opinion polarity) is more similar to TREC (question type)
	\item Sample size: We vary the sample size of the source task from $100$ to $3000$.
	\item Covariate shift: We implement the covariance alignment procedure following \cite{WZR20}.
	We fix the sample size of the target task as $1000$.
\squishend

@@

\textbf{Further results of the covariance alignment procedure.}
Our results in Figure \ref{fig_ab_cov} are averaged over all the task pairs.
In Figure \ref{fig_covariate_app}, we show two task pairs as examples.
In Figure \ref{fig_cov_a}, we observe that for the particular task pair, covariance alignment provides more significant gains when the sample ratio is large.
In Figure \ref{fig_cov_b}, we observe that covariance alignment does not always improve over the baseline multi-task learning model.
One explanation is that MR and SST are similar tasks, hence adding the alignment module is unnecessary.
An interesting question is to understand when adding the alignment module benefits the multi-task learning model.
We leave this question for future work.
%Note: For text classification tasks, the source task training data size ranges from 500 to 1,500 and target task training data size is 1000; For ChestX-ray14,

\begin{figure}[!h]
	\centering
	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\includegraphics[width=0.7\textwidth]{figures/ratio_alignment_norm_trec_cr_lstm.pdf}
		\caption{Task pair TREC and CR}
		\label{fig_cov_a}
	\end{subfigure}\hfill
		\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\includegraphics[width=0.7\textwidth]{figures/ratio_alignment_mr_sst_lstm.pdf}
		\caption{Task pair MR and SST}
			\label{fig_cov_b}
	\end{subfigure}
	\caption{(a) For the task pair TREC and CR, adding the covariance alignment procedure provides more improvement when the source/target sample ratio is large.
	(b) For the task pair MR and SST, adding the covariance alignment procedure hurts performance.
	One explanation is that MR and SST are similar tasks, hence adding the alignment module is unnecessary.}
	\label{fig_covariate_app}
\end{figure}

### 2020-08-09 arxiv/proof_main.tex
author response

\begin{lemma}[Variance bound: Restate of Lemma \ref{lem_cov_shift_informal}]\label{lem_cov_shift}
	%Let $X_i\in\real^{n_i\times p}$ be a random matrix that contains i.i.d. row vectors with mean $0$ and variance $\Sigma_i\in\real^{p\times p}$, for $i = 1, 2$.
	Suppose $X_1=Z_1\Sigma_1^{1/2}\in \R^{n_1\times p}$ and $X_2=Z_2\Sigma_2^{1/2}\in \R^{n_2\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho_1:=n_1/p>1$ and $\rho_2:=n_2/p>1$ being fixed constants.
	Denote by $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and let $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M$ in descending order. Let $A$ be any $p\times p$ matrix that is independent of $X_1$ and $X_2$. We have that for any constant $\e>0$,
	%When $n_1 = c_1 p$ and $n_2 = c_2 p$, we have that with high probability over the randomness of $X_1$ and $X_2$, the following equation holds
	\begin{align}\label{lem_cov_shift_eq}
		\bigtr{(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}A} = \frac{1}{\rho_1+\rho_2}\frac1p\bigtr{ (a_1 \Sigma_1 + a_2\Sigma_2)^{-1} A} +\bigo{\|A\| p^{-1/2+\epsilon}}
	\end{align}
with high probability, where $(a_1, a_2)$ is the solution to the following deterministic equations:
	\begin{align}
		a_1 + a_2 = 1- \frac{1}{\rho_1 + \rho_2},\quad a_1 + \frac1{\rho_1 + \rho_2}\cdot \frac{1}{p}\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2} = \frac{\rho_1}{\rho_1 + \rho_2}. \label{eq_a12extra}
	\end{align}
\end{lemma}

@@

\begin{theorem}[Restate of Theorem \ref{thm_main_informal}]\label{thm_model_shift}
%For $i=1,2$, let $Y_i = X_i\beta_i + \varepsilon_i$ be two independent data models, where $X_i$, $\beta_i$ and $\varepsilon_i$ are also independent of each other. Suppose that $X_i=Z_i\Sigma_i^{1/2}\in \R^{n_i\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho_i:=n_i/p>1$ being fixed constants, and $\e_i\in \R^{n_i}$ are random vectors with i.i.d. entries with mean zero, variance $\sigma^2$ and all moments as in \eqref{assmAhigh}. 
Consider two data models $Y_i = X_i\beta_i + \varepsilon_i$, $i=1,2$, that satisfy Assumption \ref{assm_secA2} with $\sigma_1^2=\sigma_2^2=\sigma^2$. Then with high probability, we have
	\begin{align}
	 	\te(\hat{\beta}_{t}^{\MTL}) \le \te(\hat{\beta}_t^{\STL}) \quad \text{ when: } \ \ &\Delta_{\vari} - \Delta_{\bias} \ge   \delta(\hat v) \label{upper}\\
		\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL}) \quad \text{ when: } \ \ &\Delta_{\vari} - \Delta_{\bias} \le - \delta(\hat v), \label{lower}
	\end{align}
	where
	\begin{align} %\bigtr{{\Sigma_2^{-1}}}
		\Delta_{\vari} &\define {\sigma^2}\bigbrace{\frac{1}{\rho_2 - 1} -  \frac{1}{\rho_1 + \rho_2}\cdot \frac1p \bigtr{(\hat a_1 \hat v^2 M^{\top}M + \hat a_2\id)^{-1}} } \label{Deltavarv} \\
		\Delta_{\bias} &\define (\beta_1 - \hat{v}\beta_2)^{\top} \Sigma_1^{1/2} \Pi (\hat v)\Sigma_1^{1/2} (\beta_1 - \hat{v}\beta_2). \label{Deltabetav}
	\end{align}
\end{theorem}

>>>>

\begin{lemma}[Variance bound: Lemma \ref{lem_cov_shift_informal} restated]\label{lem_cov_shift}
	%Let $X_i\in\real^{n_i\times p}$ be a random matrix that contains i.i.d. row vectors with mean $0$ and variance $\Sigma_i\in\real^{p\times p}$, for $i = 1, 2$.
	Suppose $X_1=Z_1\Sigma_1^{1/2}\in \R^{n_1\times p}$ and $X_2=Z_2\Sigma_2^{1/2}\in \R^{n_2\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho_1:=n_1/p>1$ and $\rho_2:=n_2/p>1$ being fixed constants.
	Denote by $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and let $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M$ in descending order. Let $A$ be any $p\times p$ matrix that is independent of $X_1$ and $X_2$. We have that for any constant $\e>0$,
	%When $n_1 = c_1 p$ and $n_2 = c_2 p$, we have that with high probability over the randomness of $X_1$ and $X_2$, the following equation holds
	\begin{align}\label{lem_cov_shift_eq}
		\bigtr{(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}A} = \frac{1}{\rho_1+\rho_2}\frac1p\bigtr{ (a_1 \Sigma_1 + a_2\Sigma_2)^{-1} A} +\bigo{\|A\| p^{-1/2+\epsilon}}
	\end{align}
with high probability, where $(a_1, a_2)$ is the solution to the following deterministic equations:
	\begin{align}
		a_1 + a_2 = 1- \frac{1}{\rho_1 + \rho_2},\quad a_1 + \frac1{\rho_1 + \rho_2}\cdot \frac{1}{p}\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2} = \frac{\rho_1}{\rho_1 + \rho_2}. \label{eq_a12extra}
	\end{align}
\end{lemma}

@@

\begin{theorem}[Theorem \ref{thm_main_informal} restated]\label{thm_model_shift}
%For $i=1,2$, let $Y_i = X_i\beta_i + \varepsilon_i$ be two independent data models, where $X_i$, $\beta_i$ and $\varepsilon_i$ are also independent of each other. Suppose that $X_i=Z_i\Sigma_i^{1/2}\in \R^{n_i\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho_i:=n_i/p>1$ being fixed constants, and $\e_i\in \R^{n_i}$ are random vectors with i.i.d. entries with mean zero, variance $\sigma^2$ and all moments as in \eqref{assmAhigh}. 
Consider two data models $Y_i = X_i\beta_i + \varepsilon_i$, $i=1,2$, that satisfy Assumption \ref{assm_secA2} with $\sigma_1^2=\sigma_2^2=\sigma^2$. Then with high probability, we have
	\begin{align}
	 	\te(\hat{\beta}_{t}^{\MTL}) \le \te(\hat{\beta}_t^{\STL}) \quad \text{ when: } \ \ &\Delta_{\vari} - \Delta_{\bias} \ge   \delta(\hat v) \label{upper}\\
		\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL}) \quad \text{ when: } \ \ &\Delta_{\vari} - \Delta_{\bias} \le - \delta(\hat v), \label{lower}
	\end{align}
	where
	\begin{align} %\bigtr{{\Sigma_2^{-1}}}
		\Delta_{\vari} &\define {\sigma^2}\bigbrace{\frac{1}{\rho_2 - 1} -  \frac{1}{\rho_1 + \rho_2}\cdot \frac1p \bigtr{(\hat a_1 \hat v^2 M^{\top}M + \hat a_2\id)^{-1}} } \label{Deltavarv} \\
		\Delta_{\bias} &\define (\beta_1 - \hat{v}\beta_2)^{\top} \Sigma_1^{1/2} \Pi (\hat v)\Sigma_1^{1/2} (\beta_1 - \hat{v}\beta_2). \label{Deltabetav}
	\end{align}
\end{theorem}
